---
title: "Assignment 1: reading and inspecting data with readr and dplyr"
subtitle: "ØKA2015 Applied Data Science, autumn 2026"
author: "YOUR NAME"
date: "DATE"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, comment = "#>")
```

<!-- How to use this file
     1. Rename it YourName_Assignment1.Rmd and put your name and the date above.
     2. Under each problem heading there is a code chunk. Put your commands in it,
        with a comment on every line. Run a chunk with the green play button.
     3. Write your explanation or answer as ordinary text under the chunk.
     4. Press Knit. Check that every problem shows the code and its output.
     If PDF knitting fails, change "output: pdf_document" to "output: html_document".
     Lines like this one, between <!-- and -->, are notes to you and do not appear
     in the knitted document. Delete them when you are done. -->

# Problem 1. Loading the data

```{r}
library(tidyverse)                                     # load the tidyverse
# housing <- read_csv("data/housing.csv")              # read the file; remove the # and complete the problem
```

Your comment on what you see.

# Problem 2. Your environment

```{r}

```

Which objects does your session contain, and which one holds the data?

# Problem 3. Size

```{r}

```

State the number of observations and variables in a sentence.

# Problem 4. Structure and summary

```{r}

```

Which column contains text rather than numbers, and which column has missing values?

# Problem 5. Selecting columns

```{r}

```

# Problem 6. Creating a column

```{r}

```

# Problem 7. What is dplyr for?

Write about 100 words. Name five dplyr functions and, in one line each, what they do.

# Problem 8. What is data wrangling?

Write about 100 words, with an example from this dataset.

# Use of AI tools

Delete this section if you did not use any. Otherwise, one or two lines: which tool,
and what you used it for.
