---
title: "Lecture 3: Introduction to the Tidyverse"
subtitle: "ØKA2015 Applied Data Science, University of Inland Norway"
author: "Touseef Hameed"
date: "16 September 2026"
output:
  pdf_document:
    toc: true
    toc_depth: 2
    number_sections: true
    highlight: tango
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
    highlight: tango
fontsize: 11pt
geometry: margin=2.2cm
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, comment = "#>")
```

# What this document is

This is an **R Markdown** file. It combines ordinary text with chunks of R code. When
you *knit* it, R runs every chunk and places the results into a PDF or a web page. It is
the form your assignments take: the reader sees your code, its output and your
explanation together.

The same code, without the surrounding text, is in `lecture03_tidyverse.R`. Open that
file in RStudio and run it line by line with **Ctrl + Enter**. Every line there carries a
comment explaining what it does.

# R as a calculator

R evaluates whatever you type at the console.

```{r}
1 + 1               # add
1 / 200 * 30        # divide, then multiply
(59 + 73 + 2) / 3   # parentheses first: an average
```

The `[1]` at the start of each answer indicates that the line of output begins at
element 1 of the result. It becomes meaningful once results contain several elements.

# Objects: the assignment operator `<-`

An **object** is a named location in R's memory that holds a value. The operator `<-`
places a value into it. Every assignment has the form `object_name <- value`; read it
as *"object name gets value"*. In RStudio the shortcut for `<-` is **Alt + minus**.

The running example is the coffee bar from Lecture 2, where a latte costs 52 NOK.

```{r}
cups <- 18           # store 18 under the name cups; nothing is printed
cups                 # the name alone displays the value
price <- 52          # the price of one latte, in NOK
cups * price         # revenue for the day
```

# Vectors: `c()`

A **vector** holds several values of the same kind in one object. `c()` means *combine*.
R applies an operation to every element of a vector at once, which is called vectorised
computation; it is why R code rarely needs a loop.

```{r}
cups <- c(18, 22, 27)      # Monday, Tuesday, Wednesday
cups * price               # the price applied to every element
sum(cups)                  # total cups over three days
mean(cups)                 # average per day
```

Text is written in quotation marks; numbers are not.

```{r}
days <- c("Mon", "Tue", "Wed")   # a vector of text
days
```

# Functions and arguments

A **function** is a named command: `function_name(argument1 = value1, argument2 = value2)`.
Arguments sit inside the parentheses, separated by commas. The first few may be given
by position; later ones are clearer by name.

```{r}
seq(from = 1, to = 10)               # arguments by name
seq(1, 10)                           # the same, by position
round(mean(cups), 1)                 # a function inside a function
mean(c(18, NA, 27), na.rm = TRUE)    # NA is a missing value; na.rm = TRUE ignores it
```

# Comments and object names

Everything after `#` on a line is a **comment**: R ignores it, and it explains the code to
a human reader. In your assignments, a comment above every command is required.

A name must start with a letter and may contain letters, numbers, `_` and `.` only. Use
descriptive names in one consistent style; this course uses `snake_case`.

```{r}
cups_monday <- 18          # snake_case, recommended
cupsMonday  <- 18          # camelCase, also valid
```

# Precision: R does exactly what you type

1. **R is case-sensitive.** `cups`, `Cups` and `CUPS` are three different names. The
   error message always names the object it could not find.
2. **Text requires quotation marks; numbers do not.** Writing `town <- Lillehammer`
   makes R search for an *object* called `Lillehammer`, and fail.
3. **Quotation marks and parentheses come in pairs.** If the console shows `+` and
   waits, a closing `"` or `)` is missing. Complete it, or press **Esc**.

# Packages: install once, load every session

A **package** is a collection of functions, data and documentation written by others.
The **tidyverse** is the collection of packages this course is built on. Two commands,
and they are not interchangeable:

- `install.packages("tidyverse")` **downloads** it to your laptop. Once per computer.
  With quotation marks.
- `library(tidyverse)` **loads** it for the current session. Every time you open R.
  Without quotation marks.

```{r install, eval=FALSE}
install.packages("tidyverse")   # run once on each computer you use
```

The tidyverse provides `dplyr` (which is where `glimpse()` comes from), `readr`
(`read_csv()`), `tibble`, and six more packages.

```{r}
library(tidyverse)               # load the nine core packages for this session
```

When it loads, R lists nine core packages. A message about *conflicts* is information,
not an error: two packages define a function with the same name, and the tidyverse
version takes precedence. `tidyverse_update()` reports whether newer versions exist.

If you ever see `there is no package called 'x'`, it means `install.packages("x")` has
not been run on that computer.

# Reading a CSV file with `readr`

`read_csv()` reads a comma-separated file and returns a **tibble**, the tidyverse's data
table. A tibble prints its dimensions, the type of every column, and only the first ten
rows, so a large dataset remains readable.

The example file has six students and five columns, from *R for Data Science*. There
are three routes into R; the first argument, the path, is the most important one.

**Route 1, from the web.** Nothing to download first.

```{r, eval=FALSE}
students <- read_csv("https://pos.it/r4ds-students-csv")   # the path is a web address
```

**Route 2, from a file on your laptop.** The path is relative to your *working
directory*, the folder R reads from and writes to. `getwd()` reports it. This document
uses route 2 so that it knits without an internet connection.

```{r}
students <- read_csv("data/students.csv")   # relative to the working directory
```

**Route 3, the Import button.** Environment pane, *Import Dataset > From Text (readr)*,
choose the file, and RStudio writes route 2 for you. Copy the command it shows into your
script so that the import is reproducible.

# A first look at the dataset

Always inspect before you analyse. These commands are Assignment 1, Problems 1 to 4.

```{r}
students             # dimensions, column names, column types
head(students)       # first six rows
nrow(students)       # observations
ncol(students)       # variables
names(students)      # column names
glimpse(students)    # every column on its own line, with its type
summary(students)    # basic statistics per column
```

Note the type under each column name: `<chr>` is text, `<dbl>` is a number. `AGE` was
read as text. That is a clue.

# Real data is rarely clean

Two features of this file are typical of real data.

**"N/A" is text, not a missing value.** R treats only a genuinely empty cell as missing
(`NA`). The string `"N/A"` is three characters of text. State which strings mean missing:

```{r}
students <- read_csv("data/students.csv", na = c("N/A", ""))   # na = : a named argument
students$favourite.food       # $ selects one column; row 3 is now NA
sum(is.na(students))          # is.na() is TRUE where missing; sum() counts them
colSums(is.na(students))      # per column
```

**"five" is a word, not a number.** One age is written out, so the whole column became
text. Replace the word and convert the column:

```{r}
students$AGE <- gsub("five", "5", students$AGE)   # global substitution
students$AGE <- as.numeric(students$AGE)          # text to numbers
students$AGE
mean(students$AGE, na.rm = TRUE)                  # ignoring the missing value
```

Lecture 4 performs this kind of correction with the dplyr verbs. This is a preview of
why data wrangling takes 50 to 80 per cent of a data scientist's time.

# Saving a copy, and other file types

`write_csv()` is the reverse of `read_csv()`: the object, then the path.

```{r}
dir.create("data", showWarnings = FALSE)          # make the folder if it does not exist
write_csv(students, "data/students_clean.csv")    # write the tibble to disk
file.exists("data/students_clean.csv")            # TRUE if the file is there
```

Several files can be read in one call, with `id = "file"` recording the source of each
row:

```{r, eval=FALSE}
sales_files <- c("https://pos.it/r4ds-01-sales",
                 "https://pos.it/r4ds-02-sales",
                 "https://pos.it/r4ds-03-sales")
sales <- read_csv(sales_files, id = "file")
```

Other formats use the same idea with a different function: `read_csv2()` for
semicolon-separated files, which is common in Norway because the comma is the decimal
marker; `read_tsv()` for tab-separated files; `read_delim()` for any delimiter;
`readxl::read_excel()` for Excel workbooks.

# What you can do now

- store values in objects and combine them with `c()`
- call a function with named or positional arguments
- install and load a package
- read a CSV from the web or from disk
- inspect its dimensions, names, types and summary
- recognise the two most common ways real data arrives incorrect

**Lecture 4, Friday 18 September:** the dplyr verbs `filter()`, `select()`, `mutate()`,
`arrange()`, `group_by()` and `summarise()`. That is Assignment 1, Problems 5 and 6.

# Sources

- Wickham, Çetinkaya-Rundel and Grolemund (2023), *R for Data Science*, 2nd ed.
  Introduction (*Prerequisites*), Ch. 2 *Workflow: basics*, Ch. 7 *Data import*.
  Free at <https://r4ds.hadley.nz>.
- The tidyverse: <https://www.tidyverse.org>
- readr: <https://readr.tidyverse.org>
