---
title: "How to write an R Markdown file"
subtitle: "A short guide, using lecture03_tidyverse.Rmd as the example"
author: "Touseef Hameed, ØKA2015 Applied Data Science"
date: "September 2026"
output:
  pdf_document:
    toc: true
    number_sections: true
    highlight: tango
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
fontsize: 11pt
geometry: margin=2.2cm
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>", warning = FALSE)
```

# What an R Markdown file is

An R Markdown file (`.Rmd`) is a text file that mixes three things:

1. a **header** at the top, between two lines of `---`, that says what the document
   is called and what it should become (PDF or web page);
2. **text**, written in Markdown, a plain way of marking headings, bold and lists;
3. **code chunks**, blocks of R code that run when the document is *knitted*, with their
   output placed right after them.

Pressing **Knit** in RStudio turns the file into a finished document: your explanation,
your code and its results, together. That is why assignments should use it: the reader
sees exactly what you ran and what came out.

This guide is itself an R Markdown file. Open `rmarkdown_guide.Rmd` next to the PDF and
compare them; every part described below is visible in the source.

# Create one

In RStudio: **File → New File → R Markdown…** Give it a title and your name, choose
**PDF** or **HTML**, click **OK**. RStudio opens a file with example content. Delete
everything below the header and write your own.

Save it with **Ctrl + S**, in a folder of its own, with any data in a `data/` subfolder
next to it. When knitting, R uses the folder of the `.Rmd` file as its working
directory, so `read_csv("data/students.csv")` finds the file with no `setwd()`.

# The three parts, using the lecture file

## The header

The top of `lecture03_tidyverse.Rmd` looks like this:

````markdown
---
title: "Lecture 3: Introduction to the Tidyverse"
author: "Touseef Hameed"
date: "16 September 2026"
output: pdf_document
---
````

Change the title, author and date. `output: pdf_document` makes a PDF;
`output: html_document` makes a web page. Keep the indentation exactly as shown; the
header is sensitive to spaces.

## Text

Ordinary sentences are just typed. A few marks give structure:

````markdown
# A main heading
## A smaller heading

Plain paragraph. **Bold**, *italic*, and `code in the text`.

- a bullet point
- another one

1. a numbered step
2. the next step

A link: <https://r4ds.hadley.nz>
````

In an assignment, use one `#` heading per problem, then the code, then your
explanation.

## Code chunks

A chunk starts with ```` ```{r} ```` on its own line and ends with ```` ``` ````:

````{verbatim}
```{r}
library(tidyverse)                      # load the packages
students <- read_csv("data/students.csv")   # read the data
head(students)                          # first six rows
```
````

Insert a new chunk with **Ctrl + Alt + I**. Run one chunk without knitting by clicking
the green **play** button at its top right. Its output appears under it in the editor.

When the document is knitted, that chunk produces this (the long message that
`library(tidyverse)` prints is hidden here with `message=FALSE`, see Section 4):

```{r, message=FALSE}
library(tidyverse)
students <- read_csv("data/students.csv", show_col_types = FALSE)
head(students)
```

# Chunk options you will need

Options go inside the braces after `r`, separated by commas.

| option | effect | use it for |
|---|---|---|
| `eval=FALSE` | show the code, do not run it | `install.packages("tidyverse")`, which must not run at every knit |
| `echo=FALSE` | run the code, hide it, show only the output | a plot where the code is not the point |
| `include=FALSE` | run the code, show nothing | a setup chunk |
| `message=FALSE, warning=FALSE` | hide messages and warnings | the long message that `library(tidyverse)` prints |

Example, as written in the lecture file:

````{verbatim}
```{r, eval=FALSE}
install.packages("tidyverse")   # run once on each computer you use
```
````

# Knit

Click **Knit** (the ball-of-wool icon above the editor), or press **Ctrl + Shift + K**.
The document is built in a **fresh R session**: nothing from your console is available.
Every package must be loaded and every object created *inside* the document, in order.

The result is saved next to the `.Rmd` file, with the same name and `.pdf` or `.html`.

PDF needs a LaTeX installation. If knitting to PDF fails, either run once in the
console

```{r, eval=FALSE}
install.packages("tinytex")
tinytex::install_tinytex()
```

or change the header to `output: html_document`. HTML is accepted for the assignment.

# The errors everyone meets

| message | cause | fix |
|---|---|---|
| `object 'students' not found` | the chunk that creates it is missing, or comes later in the file | create every object before you use it, inside the document |
| `could not find function "read_csv"` | `library(tidyverse)` is not in the document, or comes after the chunk | put it in the first chunk |
| `cannot open file 'data/students.csv'` | the file is not in a `data` folder next to the `.Rmd` | move it there, or use the web address |
| the knit stops at `install.packages` | the chunk is being run | add `eval=FALSE` to that chunk |
| `Error in parse: unexpected symbol` | a typing error in the chunk, often a missing comma or quote | run the chunk alone with the play button to find the line |

The knitting log, in the **R Markdown** tab at the bottom of RStudio, names the chunk
where the problem is.

# For the assignment

- One heading per problem: `# Problem 1`, `# Problem 2`, and so on.
- Under each: the chunk with the code and, for the written problems, your text.
- A comment on every command.
- Knit it, check that the output under each chunk is what you meant, and upload the
  `.Rmd` together with the knitted `.pdf` or `.html`.

If knitting does not work on your computer, screenshots of your RStudio window showing
the code and its output are accepted instead.
