---
title: "Practice exercise, Lecture 7"
subtitle: "ØKA2015 Applied Data Science. Not graded. Do it before the next lecture."
author: "Touseef Hameed"
date: "6 October 2026"
output:
  pdf_document: default
  html_document: default
fontsize: 11pt
geometry: margin=2.2cm
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

Reference: the Lecture 7 slides and `lecture07_descriptive.R`. You use the same steps on
a new table: the California housing data from Assignment 1, read straight from the web:

```{r}
library(tidyverse)
url <- paste0("https://raw.githubusercontent.com/touseefhameed/",
              "oka2015-applied-data-science/main/assignment1/data/housing.csv")
housing <- read_csv(url)
```

`paste0()` glues the two pieces of text into one web address; it is split only so that
it fits on the page. From your Assignment 1 project, `read_csv("data/housing.csv")` does
the same.

Each row is a district in California. Work in a script, and write one sentence under each
answer. Eight tasks, about 50 minutes.

## Task 1: types of data

a. Use `glimpse()`. Which columns are quantitative and which are categorical?
b. Is `housing_median_age` discrete or continuous? Is `ocean_proximity` nominal or ordinal?
c. Is this cross-sectional or time series data?

## Task 2: a frequency table

Make a frequency table of `ocean_proximity` with a percent column, largest first. Which
category is smallest, and why might that matter later?

## Task 3: shape

Draw a histogram of `median_income` and one of `housing_median_age`. Describe each shape.
Check your answer by comparing the mean and the median.

## Task 4: centre

a. The mean, median and mode of `median_house_value`.
b. The same three measures for each `ocean_proximity` group, with `n`. Which group has
   the highest median value?

## Task 5: spread

a. The standard deviation, the range and the IQR of `median_house_value`.
b. Use `summary()` on the same column. What does the maximum tell you?

## Task 6: missing values

a. Which column has missing values, and how many?
b. Look at its shape. Would you fill the missing values with the mean or the median? Fill
   them, and check that none are left.

## Task 7: outliers

a. Draw a boxplot of `median_house_value`. Count the districts outside the IQR fences.
b. Count the districts with a value of exactly 500,001. Is that a real price or a limit
   set by whoever collected the data? What would you do with them?
c. Use z-scores to count the districts whose `total_rooms` is more than 3 sd from the mean.

## Task 8: association

a. Draw a scatter plot of `median_income` against `median_house_value`.
b. The correlation between them. Strong or weak? Positive or negative?
c. The correlation matrix of `median_income`, `median_house_value`, `housing_median_age`
   and `population`. Which pair is the weakest?
d. Does a high correlation between income and house value prove that income causes high
   prices? Why not?

---

## Answers to "Check yourself" from the script

1. `ocean_proximity` is categorical. Its categories have no natural order, so it is
   nominal.
2. Mean above median: skewed right. A few very expensive houses pull the mean up.
3. The standard deviation is in the units of the data; the variance is in squared units.
4. IQR = 30 - 10 = 20. Fences: 10 - 1.5 × 20 = -20 and 30 + 1.5 × 20 = 60.
5. A clear downward cloud of points: as one variable rises, the other falls.

*The answers to Tasks 1 to 8 are not distributed now. Compare with a classmate, or bring
your script to the next lecture.*
