---
title: "Lecture 7: Descriptive statistics and exploratory analysis"
subtitle: "ØKA2015 Applied Data Science, University of Inland Norway"
author: "Touseef Hameed"
date: "6 October 2026"
output:
  pdf_document:
    toc: true
    toc_depth: 2
    number_sections: true
    highlight: tango
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
    highlight: tango
fontsize: 11pt
geometry: margin=2.2cm
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, comment = "#>",
                      fig.width = 5.5, fig.height = 3.4, fig.align = "center")
options(width = 80, tibble.print_max = 5, tibble.print_min = 5, pillar.sigfig = 5)
```

# What this document is

This is Lecture 7 as a written tutorial, in the same order as the script
`lecture07_descriptive.R` and as Chapter 2 of *Business Analytics* (Camm et al.).
Read it after the lecture, or run it: open the `.Rmd` file in RStudio, press **Knit**,
and this PDF is rebuilt on your computer.

In Lectures 3 to 6 you brought data into R and changed it. Today you describe it with
numbers: what kind of data it is, where its centre is, how spread out it is, which values
are unusual, and how two variables move together.

# Packages and the data

Three packages. Install each once per computer; running the install again does no harm.

```{r install, eval=FALSE}
install.packages("tidyverse")
install.packages("nycflights13")
install.packages("GGally")
```

Load them every session:

```{r packages}
library(tidyverse)
library(nycflights13)
library(GGally)
flights <- nycflights13::flights
weather <- nycflights13::weather
```

`flights` is the table from Lectures 4 to 6. `weather` has one row per hour at the three
New York airports in 2013: temperature, humidity, wind, pressure and visibility.

# Types of data

## Population and sample

The **population** is every element of interest. A **sample** is a subset of it. Here the
population is all 2013 flights from New York. `slice_sample()` draws a random sample, and
`set.seed()` fixes the random draw so that everyone gets the same sample.

```{r sample}
set.seed(2026)
flights_sample <- flights |>
  slice_sample(n = 1000)
mean(flights$dep_delay, na.rm = TRUE)
mean(flights_sample$dep_delay, na.rm = TRUE)
```

The sample mean, 12.9 minutes, is close to the population mean, 12.6. A random sample is
representative: on average it tells the truth about the population. In practice we
nearly always work with samples.

## Quantitative and categorical

A **quantitative** variable is a number you can add and average: `<int>` or `<dbl>` in R.
It is *discrete* if it counts (month, flights per day) and *continuous* if it measures
(distance, air time). A **categorical** variable is a label you can only count: `<chr>`
or `<fct>`. It is *nominal* if the categories have no order (carrier, origin) and
*ordinal* if they do (low, medium, high).

```{r types}
flights |>
  select(carrier, month, dep_delay) |>
  glimpse()
```

Careful: a number can be a label. A postcode or a 1 to 5 rating is categorical, even if
R stores it as a number.

## Cross-sectional and time series

**Cross-sectional** data describe many units at about the same point in time. A **time
series** measures one quantity over many time periods.

```{r cross-time}
flights |> filter(month == 1, day == 1) |> nrow()
flights |> count(month)
```

The 842 flights of 1 January are a cross-section. The number of flights per month is a
time series.

# Distributions

**Descriptive analysis** summarises what is in the data. **Exploratory analysis** looks for
patterns and relationships you did not know about. Exploration suggests questions; it
does not prove answers. Correlation is not causation.

A **distribution** shows which values a variable takes, and how often.

## A categorical variable: a frequency table

```{r freq}
flights |>
  count(carrier, sort = TRUE) |>
  mutate(percent = round(100 * n / sum(n), 1))
```

`n` is the frequency; `percent` is the percent frequency. United flew 17.4% of the flights.

## A quantitative variable: a histogram

A histogram splits the range into bins and draws one bar per bin. `breaks` sets the
number of bins and `xlim` the part of the x-axis shown.

```{r hist}
hist(flights$dep_delay, breaks = 200, xlim = c(-50, 300))
```

## The shape of a distribution

```{r shapes, fig.height = 2.8}
hist(weather$pressure)
hist(weather$visib)
```

Pressure is **symmetric**: the left side mirrors the right, and the mean (1017.9) is
close to the median (1017.6). Departure delay is **skewed right**: a long tail to the
right, and the mean (12.6) is above the median (-2). Visibility is **skewed left**: a long
tail to the left, and the mean (9.3) is below the median (10).

# Central tendency

## Mean, median, mode

The **mean** is the sum divided by the count. The **median** is the middle value once the
values are sorted. The **mode** is the most frequent value. R has no function for the
mode, but `count(sort = TRUE)` puts it on top.

```{r centre-toy}
books <- c(3, 7, 2, 9, 7, 5, 7)
mean(books)
median(books)
tibble(books) |> count(books, sort = TRUE)
```

## Mean or median?

```{r centre-real}
mean(flights$dep_delay, na.rm = TRUE)
median(flights$dep_delay, na.rm = TRUE)
flights |> count(dep_delay, sort = TRUE)
```

The mean uses every value, so a few very late flights pull it up to 12.6 minutes. The
median uses only the order: the typical flight left two minutes early. For skewed data,
report the median, or both.

## The centre for each group

```{r centre-group}
flights |>
  group_by(origin) |>
  summarise(
    mean   = mean(dep_delay, na.rm = TRUE),
    median = median(dep_delay, na.rm = TRUE),
    n      = n())
```

Newark has the highest mean delay. All three medians are below zero. Always report `n`
beside an average.

## Filling a missing value

The shape decides which value to fill with: the mean for a symmetric quantitative
variable, the median for a skewed one or one with outliers, the mode for a categorical
one.

```{r fill}
weather_filled <- weather |>
  mutate(pressure = replace_na(pressure, mean(pressure, na.rm = TRUE)))
sum(is.na(weather$pressure))
sum(is.na(weather_filled$pressure))
flights |> count(carrier, sort = TRUE) |> slice(1)
```

Filling changes the data. Do it only with a reason, and say so in your report.

# Variability

## Variance and standard deviation

The **variance** is the average squared distance from the mean, dividing by $n - 1$ for a
sample. The **standard deviation** is its square root, back in the units of the data.

```{r spread-toy}
scores <- c(70, 75, 80, 85, 90)
tibble(scores) |>
  mutate(deviation = scores - mean(scores),
         squared   = deviation^2)
var(scores)
sd(scores)
```

The squared deviations sum to 250, so the variance is 250 / 4 = 62.5 and the standard
deviation is 7.9 points. We report the standard deviation because it is in the units of
the data.

## Spread on real data

```{r spread-real}
sd(flights$dep_delay, na.rm = TRUE)
range(flights$dep_delay, na.rm = TRUE)
quantile(flights$dep_delay, c(0.25, 0.50, 0.75), na.rm = TRUE)
IQR(flights$dep_delay, na.rm = TRUE)
summary(flights$dep_delay)
```

The **range** uses only the two extreme values. The *p*th **percentile** has *p*% of the
values below it. The **quartiles** are the 25th, 50th and 75th percentiles. The **IQR**,
Q3 minus Q1, is the spread of the middle 50%: here 16 minutes. Extreme values do not move
it. The **coefficient of variation**, sd divided by the mean, gives the spread relative
to the mean, without units.

# Outliers

An **outlier** is an observation far from the others. It can be an error or a real but
rare case. Find it, check where it comes from, then decide. A confirmed error is corrected
or removed, and you write down what you did. A real value is kept, and you use measures
it does not distort, such as the median and the IQR.

## The IQR rule and the boxplot

Values below Q1 - 1.5 IQR or above Q3 + 1.5 IQR are flagged. For departure delay the
fences are -5 - 24 = -29 and 11 + 24 = 35.

```{r boxplot}
boxplot(flights$dep_delay, ylim = c(-50, 150))
flights |>
  filter(dep_delay < -29 | dep_delay > 35) |>
  nrow()
```

The box runs from Q1 to Q3, the thick line is the median, and the dots are flagged values.
In skewed data the rule flags many real values: 43,216 flights here. A flag is not a
verdict.

## The z-score rule

A z-score is how many standard deviations a value lies from the mean:
$z = (x - \bar{x}) / s$. For a bell-shaped distribution about 68% of values lie within 1
sd, 95% within 2 and nearly all within 3, so $|z| > 3$ is rare and worth a look.

```{r zscore}
m <- mean(flights$dep_delay, na.rm = TRUE)
s <- sd(flights$dep_delay, na.rm = TRUE)
flights |>
  mutate(z = (dep_delay - m) / s) |>
  filter(abs(z) > 3) |>
  nrow()
```

The z-score rule works best for symmetric data. For skewed data, prefer the IQR rule.

## A real data entry error

```{r wind}
weather |>
  arrange(desc(wind_speed)) |>
  select(origin, month, day, hour, wind_speed)
```

A hurricane is about 75 mph, so 1,048 mph is a recording error. Remove it, and write down
that you did:

```{r wind-clean}
weather_clean <- weather |>
  filter(wind_speed < 200 | is.na(wind_speed))
nrow(weather) - nrow(weather_clean)
```

The longest departure delay, 1,301 minutes on 9 January, is a real flight. Keep it.

# Association between two variables

## Scatter plots

```{r scatter, fig.width = 7, fig.height = 2.6}
par(mfrow = c(1, 3))
plot(flights_sample$distance, flights_sample$air_time)
plot(flights_sample$dep_delay, flights_sample$arr_delay)
plot(flights_sample$distance, flights_sample$dep_delay)
```

Longer flights take longer: a tight upward line. Late departures arrive late: upward.
Distance says nothing about delay: no pattern. (`par(mfrow = c(1, 3))` puts the three
plots side by side in this document; in RStudio you can run the three `plot()` lines one
at a time.)

## Covariance and correlation

```{r cov-cor}
cov(flights$distance, flights$air_time, use = "complete.obs")
cor(flights$distance, flights$air_time, use = "complete.obs")
cor(flights$dep_delay, flights$arr_delay, use = "complete.obs")
cor(flights$distance, flights$dep_delay, use = "complete.obs")
```

**Covariance** is positive when two variables move together and negative when they move
apart, but its size depends on the units. **Correlation** rescales it to lie between -1
and +1, with no units: +1 is a perfect positive relationship, 0 no linear relationship,
-1 a perfect negative one. `use = "complete.obs"` skips rows with a missing value.

## Many pairs at once

```{r cor-matrix}
flights |>
  select(dep_delay, arr_delay, air_time, distance) |>
  cor(use = "complete.obs") |>
  round(2)
```

```{r ggpairs, fig.width = 6, fig.height = 4.4}
flights_sample |>
  select(dep_delay, arr_delay, distance) |>
  ggpairs()
```

Correlation is not causation. Late departures and late arrivals move together
($r = 0.91$), but these numbers alone cannot say why.

# What you can do now

- name the type of a variable and of a dataset
- describe a distribution: frequency table, histogram, shape
- report the centre (mean, median, mode) and the spread (sd, IQR)
- flag outliers with the IQR rule and z-scores, and decide what to do
- describe two variables together with a scatter plot and a correlation

# Sources

- Camm, J. D., Fry, M. J., Cochran, J. J. and Ohlmann, J. W. (2024). *Business Analytics*.
  Cengage. Ch. 2, Descriptive Statistics.
- Wright, C., Ellis, S., Hicks, S. and Peng, R. (2021). *Tidyverse Skills for Data
  Science in R*, Ch. 5.4 and 5.5. <https://jhudatascience.org/tidyversecourse/>
- Course repository: <https://github.com/touseefhameed/oka2015-applied-data-science>
