---
title: "Lecture 4: Data Lab, importing and transforming data with dplyr (rows)"
subtitle: "ØKA2015 Applied Data Science, University of Inland Norway"
author: "Touseef Hameed"
date: "18 September 2026"
output:
  pdf_document:
    toc: true
    toc_depth: 2
    number_sections: true
    highlight: tango
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
    highlight: tango
fontsize: 11pt
geometry: margin=2.2cm
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, comment = "#>")
options(width = 80, tibble.print_max = 3, tibble.print_min = 3, pillar.sigfig = 4)
```

# What this document is

This is the Lecture 4 Data Lab as a written tutorial, in the same order as the script
`lecture04_wrangling.R` and as Chapter 3 of *R for Data Science* up to the end of
"Rows". The explanation is text between the code instead of comments beside it. Read it
after the lab, or run it: open the `.Rmd` file in RStudio, press **Knit**, and this PDF
is rebuilt on your computer.

Lecture 3 brought a dataset into R and looked at it. Today you start changing it. You
rarely get data in exactly the form you need; you keep some rows, reorder them, and
count what is there. Those are the *row verbs* of the **dplyr** package. Columns and
groups follow in the next lectures.

# Packages and the data

Two packages: the **tidyverse**, which provides dplyr (the verbs, and `glimpse()`) and
readr (`read_csv()`), and **nycflights13**, which contains the data. Install each once
per computer. Running the install again does no harm; it only takes time.

```{r install, eval=FALSE}
install.packages("tidyverse")
install.packages("nycflights13")
```

Then load both, every session:

```{r packages}
library(tidyverse)
library(nycflights13)
flights <- nycflights13::flights
```

`library(tidyverse)` prints a *Conflicts* note: `dplyr::filter()` masks
`stats::filter()`. Two packages have a function called `filter`, and the tidyverse one
is used. The double colon, `packagename::function()`, says precisely which one you mean.
`nycflights13::flights` reads as "flights from the package nycflights13".

On Wednesday the data came from a CSV file. A package that ships its data is a fourth
way to get data into R. Everything that follows is the same whichever way you used.

# Quick inspection: look before you touch

`flights` is a tibble: 336,776 flights that left New York in 2013, 19 columns. A tibble
prints only the first rows and the columns that fit. Four ways to see more:
`View(flights)` opens the spreadsheet viewer in RStudio, `print(flights, width = Inf)`
prints every column, `print(flights, n = Inf)` prints every row (all 336,776, so use it
with care), and `glimpse()` lists every column with its type.

```{r inspect}
flights
nrow(flights)
ncol(flights)
glimpse(flights)
```

The type under each name decides what you can do with it: `<int>` and `<dbl>` are
numbers, `<chr>` is text, `<dttm>` a date-time. Delays are in minutes, negative means
early, and `dep_time` 517 means 05:17. `NA` in `dep_delay` means the flight was
cancelled.

`$` pulls one column out as a plain vector, which goes straight into `mean()`, `sum()`
or `max()`:

```{r dollar}
head(flights$dep_delay)
mean(flights$distance)
```

# Operators, the pipe, and `na.rm`

A condition compares values and gives TRUE or FALSE. Two equals signs ask "is it
equal?". Text is compared exactly, capitals included.

```{r compare}
3 == 3
3 != 4
5 >= 2
"JFK" == "jfk"
```

Logical operators combine TRUE and FALSE: `&` is AND, `|` is OR, `!` is NOT. `%in%`
asks whether a value is one of a set, and `a:b` is a range.

```{r logical}
TRUE & FALSE
TRUE | FALSE
!TRUE
"JFK" %in% c("JFK", "LGA", "EWR")
7:9
```

**The pipe.** `x |> f(y)` means `f(x, y)`: take `x`, *then* apply `f`. Read `|>` as
"then". **Ctrl + Shift + M** inserts it in RStudio; if that gives `%>%`, the older
pipe, tick *Use native pipe operator* under Tools, Global Options, Code. The two behave
the same for everything in this course.

```{r pipe}
sqrt(16)
16 |> sqrt()
flights |> nrow()
```

The pipe is useful when there are several steps, because each line reads as a step in a
sentence: take the flights, then keep January, then average the departure delay.

```{r pipe-chain}
flights |>
  filter(month == 1) |>
  summarise(avg_delay = mean(dep_delay))
```

The answer is `NA`. `mean()`, `sum()` and `max()` return `NA` if any value is missing,
and cancelled flights have no departure delay. `na.rm = TRUE` tells the function to skip
the missing values for this one calculation; the data itself is unchanged.

```{r narm}
v <- c(1, 2, NA, 4)
mean(v)
mean(v, na.rm = TRUE)
flights |>
  filter(month == 1) |>
  summarise(avg_delay = mean(dep_delay, na.rm = TRUE))
```

Note the two different signs in that code. Inside `summarise()`, `=` gives the result a
name, `avg_delay`. Outside a function call, `<-` stores an object, as in `v <- ...`.
They are not interchangeable: `=` names something inside a call, `<-` creates an object.

**What every dplyr verb shares.** The first argument is a data frame; the pipe supplies
it. The next arguments name columns, without quotes. The result is a new data frame,
and the input is never changed. That is why these two lines are the same command. Here
`|> nrow()` is added so that only the row count prints; without it you get the whole
table.

```{r shares}
filter(flights, month == 1) |> nrow()
flights |> filter(month == 1) |> nrow()
```

# Rows: `filter()`

`filter()` keeps the rows for which a condition is TRUE. All the columns stay.

```{r filter}
flights |> filter(dep_delay > 120)
```

9,723 of 336,776 flights left more than two hours late. Conditions combine with `&`
(and), `|` (or) and `%in%` (is one of). Only the row counts are printed below; each
line returns a whole table like the one above.

```{r filter-more}
flights |> filter(month == 1 & day == 1) |> nrow()
flights |> filter(month == 1 | month == 2) |> nrow()
flights |> filter(month %in% c(1, 2)) |> nrow()
flights |> filter(dest == "IAH") |> nrow()
```

January and the first: New Year's Day, 842 flights. January or February, written two
ways: 51,955 flights. Houston, `"IAH"` in quotes because it is text: 7,198 flights.

`is.na()` is a condition like any other: TRUE where a value is missing. With `!` in
front it keeps the rows that have a value.

```{r filter-na}
flights |> filter(is.na(dep_delay)) |> nrow()
flights |> filter(!is.na(dep_delay)) |> nrow()
```

8,255 cancelled flights, and 328,521 that departed.

Two common mistakes. The first stops with a helpful message:

```{r mistake-1, error=TRUE}
flights |> filter(month = 1)
```

The second gives no error and is wrong: R reads `month == 1 | 2` as
`(month == 1) | 2`, and the number 2 counts as TRUE, so every row passes.

```{r mistake-2}
flights |> filter(month == 1 | 2) |> nrow()
```

Write `month == 1 | month == 2`, or `month %in% c(1, 2)`.

A verb returns a new table and changes nothing. To keep a result, name it with `<-`,
and a second object appears in the Environment pane:

```{r keep}
jan1 <- flights |> filter(month == 1 & day == 1)
nrow(flights)
nrow(jan1)
```

# Rows: `arrange()`

`arrange()` changes the order of the rows. Ascending by default, descending with
`desc()`. The row count never changes.

```{r arrange}
flights |> arrange(dep_delay)
flights |> arrange(desc(dep_delay))
```

The longest departure delay was 1,301 minutes, on 9 January.

With several columns, the first column decides the order and each later column breaks
ties in the ones before it. `arrange(year, month, day, dep_time)` puts the earliest
year first; within a year, the earliest month (January before December); within a
month, the earliest day; within a day, the earliest departure time (05:00 before
22:00). The result is chronological order.

```{r arrange-keys}
flights |> arrange(year, month, day, dep_time) |> nrow()   # same table, chronological
flights |> arrange(desc(dep_time))
```

`desc()` wraps exactly one column. `desc(dep_time)` puts the latest departure times
first.

The printout shows only the columns that fit, so after sorting by `distance` you may
not see `distance` or `dest` at all. Pipe into `glimpse()` to see every column of the
top rows, or open the result with `View()`.

```{r arrange-glimpse}
flights |> arrange(desc(distance)) |> glimpse()
```

# Rows: `distinct()` and `count()`

`distinct()` lists each unique combination of the named columns once. `count()` adds
how many rows each has, in a column `n`; `sort = TRUE` puts the largest first.

```{r distinct}
flights |> distinct(origin, dest)
flights |> distinct(year, month, day) |> nrow()
flights |> count(origin, dest, sort = TRUE)
flights |> count(dest, sort = TRUE)
```

New York flew 224 routes in 2013, on every one of the 365 days. JFK to Los Angeles was
the busiest route and Chicago O'Hare the busiest destination. `count()` is the first
question to ask of any text column: which values occur, and are any of them typos?

# Check yourself

From the exercises of Chapter 3. The answers are at the end of the practice exercise.

1. Flights with an arrival delay of two or more hours.
2. Flights that flew to Houston, `"IAH"` or `"HOU"`.
3. Flights operated by United, American or Delta: `"UA"`, `"AA"`, `"DL"`.
4. Flights that departed in summer: July, August, September.
5. Flights that arrived more than two hours late but did not leave late.
6. Sort flights to find the longest departure delays; then the flights that left
   earliest in the morning.
7. Was there a flight on every day of 2013?
8. Which flights travelled the farthest? Which the least?

# What you can do now

- see a table: print, `View()`, `head()`, `glimpse()`, `$`
- write a condition with `== != > >= < <= & | ! %in% is.na()`
- chain steps with the pipe `|>` and read them as "then"
- rows: `filter()`, `arrange()` with `desc()`, `distinct()`, `count()`
- keep a result with `<-`, and know that a verb never changes its input

The next lectures continue with dplyr: columns, then groups.

# Sources

- Wickham, H., Çetinkaya-Rundel, M. and Grolemund, G. (2023). *R for Data Science*,
  2nd ed. Ch. 3, Data transformation, sections Introduction to Rows. Free at
  <https://r4ds.hadley.nz/data-transform>.
- Wright, C., Ellis, S., Hicks, S. and Peng, R. (2021). *Tidyverse Skills for Data
  Science in R*, Ch. 3. <https://jhudatascience.org/tidyversecourse/>
- Course repository: <https://github.com/touseefhameed/oka2015-applied-data-science>
